---
title: "Using mbd"
author: "Giovanni Laudanno"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Using mbd"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`mbd` is a package to support multiple birth model. 
This model accounts for a diversification process in which multiple 
birth (at different nodes) are allowed to occur at the same time.

The basic framework is described in Etienne et al. 2012^[
  Etienne, R.S., Haegeman, B., Stadler, T., Aze, T., Pearson, P.N., Purvis, A., 
  Phillimore, A.B.: 
  Diversity-dependence brings molecular phylogenies closer to agreement with the
  fossil record. Proceedings of the Royal Society of 
  London B: Biological Sciences
  279(1732), 1300–1309 (2012) 
].
The main purpose of the package is to provide a tool to estimate, given a phylogenetic tree, the likelihood for a set of three parameters: $\lambda$,$\mu$ and $q$.
Here:

- $\lambda$ is the sympatric speciation rate;

- $\mu$ is the extinction rate;

- $\nu$ is the multiple allopatric speciation trigger rate;

- $q$ is the single-lineage speciation probability.

Those parameters may be inferred through a maximum likelihood approach using the function mbd_ML.

## Setup

We will need to load the package:

```{r}
library(mbd)
```

Also, we'll set the random number generator seed to a value,
so that this vignette always produces the same results:

```{r}
set.seed(14)
```

## Simulating an MBD tree

First we set the parameters of the MBD speciation model:

```{r}
lambda <- 0.2 # sympatric speciation rate
mu <- 0.15 # extinction rate;
nu <- 2.0 # multiple allopatric speciation trigger rate
q <- 0.1 # single-lineage speciation probability
sim_pars <- c(lambda, mu, nu, q); 
```

Here "sim_pars" are $\lambda$, $\mu$, $\nu$ and $q$, 
i.e. the parameters we want to eventually be able to infer. 

Then we set some basic parameters to create a simulated tree:

```{r}
soc <- 2 # Use a crown age
crown_age <- 10
cond <- 1 # Condition on non-extinction

sim <- mbd_sim(
  pars = sim_pars, 
  soc = soc, 
  age = crown_age, 
  cond = cond, 
  tips_interval = c(0, Inf)
)
```

These are additional auxilary parameters 
such as `soc` (equal to 1 if we want a stem, or 2 if we want a crown), 
`cond` (let us condition the tree on the survival of crown species) 
and `crown_age` (being the total age of the tree).

The `sim` list contain different objects.
You can visualize the full tree and keep track of species that die 
before present time (missing species):

```{r, fig.show='hold', fig.width=7, fig.height=7}
graphics::plot(sim$tas)
```

If you want you can also show the reconstructed tree, 
obtained cutting the kinks from the previous tree:

```{r, fig.show='hold', fig.width=7, fig.height=7}
graphics::plot(sim$tes)
```

The dataframe containing all the information related to this tree 
is labeled as L (or L-table). You can access it by typing:

```{r}
knitr::kable(head(sim$L))
```

In this table:

- the first column are the times at which species are born;

- the second contains labels of the species' parents; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the third contains the labels of the daughter species themselves; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the fourth contains the extinction times of the species. If this is equal to -1, then the species is still extant.

To recover the vector of branching times use:

```{r}
knitr::kable(head(sim$brts))
```

In this vector values repeated more than once indicate 
that a multiple birth process has taken place.

## Likelihood estimation

To estimate the likelihood of a phylogeny, one needs

 * parameter estimates
 * the branching times (topology is irrelevant)

In this case, we'll use 

 * the parameter estimates that we used to generate the tree
 * the branching times of the tree we generated

The function `mbd_loglik` shows the likelihood of the parameters having
generated the branching times:

```{r}
mbd::mbd_loglik(
  pars = sim_pars, 
  brts = sim$brts, 
  soc = soc, 
  cond = cond, 
  missnumspec = 0
)
```

## Maximum likelihood estimation

You can also ask for an optimization of the likelihood function 
using the function `mbd_ML` either considering a variation of all 
the parameters ($\lambda, \mu, \nu, q$) or any their subset of your choice. 

Because this is a heavy calculation, we'll use a simpler tree:

```{r}
sim <- mbd_sim(
  pars = sim_pars, 
  soc = soc, 
  age = crown_age / 10, 
  cond = cond, 
  tips_interval = c(0, Inf)
)
ape::plot.phylo(sim$tes)
```

If you want for example to maximize the likelihood only for the parameter q:

```{r}
idparsopt <- 4 # Only optimize the fourth parameter, q
ids <- 1:4
idparsfix <- ids[-idparsopt] # Fix all parameters except q
parsfix <- sim_pars[idparsfix] # Use the known values for the fixed parameters
initparsopt <- 0.15 # Set an initial guess for q
mbd_ML(
  brts = sim$brts, 
  initparsopt = initparsopt, 
  idparsopt = idparsopt, 
  parsfix = parsfix, 
  idparsfix = idparsfix, 
  soc = soc, 
  cond = cond
)
```
