---
title: "Using mbd"
author: "Giovanni Laudanno and Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Using mbd"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Introduction

`mbd` is a package to support multiple birth model.
This model accounts for a diversification process in which multiple
birth (at different nodes) are allowed to occur at the same time.

The basic framework is described in Etienne et al. 2012^[
  Etienne, R.S., Haegeman, B., Stadler, T., Aze, T., Pearson, P.N., Purvis, A.,
  Phillimore, A.B.:
  Diversity-dependence brings molecular phylogenies closer to agreement with the
  fossil record. Proceedings of the Royal Society of
  London B: Biological Sciences
  279(1732), 1300–1309 (2012)
].
The main purpose of the package is to provide a tool to estimate, given a phylogenetic tree, the likelihood for a set of four parameters: $\lambda$, $\mu$, $\nu$ and $q$.
Here:

- $\lambda$ is the sympatric speciation rate;

- $\mu$ is the extinction rate;

- $\nu$ is the multiple allopatric speciation trigger rate;

- $q$ is the single-lineage speciation probability.

Those parameters may be inferred through a maximum likelihood approach using the function `mbd_ml`.

## 2. Setup

We will need to load the package:

```{r}
# devtools::install_github("Giappo/mbd", quiet = TRUE)
library(mbd)
```

Also, we will set the random number generator seed to a value,
so that this vignette always produces the same results:

```{r}
set.seed(2)
```

## 3. Simulating an MBD tree

First we set the parameters of the MBD speciation model:

```{r}
lambda <- 0.2 # sympatric speciation rate
mu <- 0.15 # extinction rate;
nu <- 2.0 # multiple allopatric speciation trigger rate
q <- 0.1 # single-lineage speciation probability
```

Then we set some basic parameters to create a simulated tree:

```{r}
if (1 == 2) {
  crown_age <- 1
  sim_pars <- c(lambda, mu, nu, q)
  sim <- mbd_sim(
    pars = sim_pars, 
    n_0 = 2, # Use a crown age 
    age = crown_age,
    cond = 1 # Condition on non-extinction
  )
}
```

These are additional auxilary parameters
such as `n_0` (equal to 1 if we want a stem, or 2 if we want a crown),
`cond` (let us condition the tree on the survival of crown species)
and `crown_age` (being the total age of the tree).

## 4. Showing the results

The `sim` list contain different objects.
You can visualize the full tree and keep track of species that die
before present time (missing species):

```{r, fig.show='hold', fig.width=7, fig.height=7}
if (1 == 2) {
  graphics::plot(sim$full_tree)
}
```

If you want you can also show the reconstructed tree,
obtained cutting the kinks from the previous tree:

```{r, fig.show='hold', fig.width=7, fig.height=7}
if (1 == 2) {
  graphics::plot(sim$reconstructed_tree)
}
```

The dataframe containing all the information related to this tree
is labeled as L (or L-table). You can access it by typing:

```{r}
if (1 == 2) {
  knitr::kable(head(sim$l_matrix))
}
```

In this table:

- the first column are the times at which species are born;

- the second contains labels of the species' parents; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the third contains the labels of the daughter species themselves; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the fourth contains the extinction times of the species. If this is equal to -1, then the species is still extant.

To recover the vector of branching times use:

```{r}
if (1 == 2) {
  knitr::kable(head(sim$brts))
}
```

In this vector values repeated more than once indicate
that a multiple birth process has taken place.

## 5. Likelihood estimation

To estimate the likelihood of a phylogeny, one needs

 * parameter estimates
 * the branching times (topology is irrelevant)

In this case, we will use

 * the parameter estimates that we used to generate the tree
 * the branching times of the tree we generated

The function `mbd_loglik` shows the likelihood of the parameters having
generated the branching times:

```{r}
if (1 == 2) {
  mbd::mbd_loglik(
    pars = c(lambda, mu, nu, q), 
    brts = sim$brts, 
    n_0 = 2, # Crown age 
    cond = 1  # Non-extinction 
  )
}
```

## 6. Maximum likelihood estimation

You can also ask for an optimization of the likelihood function
using the function mbd_ml either considering a variation of all
the parameters ($\lambda, \mu, \nu, q$) or any their subset of your choice.

Because this is a heavy calculation, we will use a simpler tree:

```{r}
phylogeny <- ape::read.tree(text = "((A:1, B:1):2, C:3);")  
ape::plot.phylo(phylogeny)
brts <- ape::branching.times(phylogeny)
```

If you want for example to maximize the likelihood only for the parameter $q$:

```{r}
if (1 == 2) {
  brts <- sim$brts
  start_pars <- c(0.2, 0.15, 1, 0.15)
  optim_ids <- c(FALSE, FALSE, FALSE, TRUE)
  n_0 <- 2
  cond <- 1
  out <- mbd::mbd_ml(
    start_pars = start_pars,
    true_pars = sim_pars,
    optim_ids = optim_ids,
    brts = brts,
    cond = cond,
    n_0 = n_0,
    verbose = TRUE
  )
  knitr::kable(out)
}
```
