---
title: "Using MBD"
author: "Giovanni Laudanno"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Using MBD"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## MBD introduction

MBD is a package to support multiple birth model. This model accounts for a diversification process in which multiple birth (at different nodes) are allowed to occur at the same time.
The basic framework is described in Etienne et al. 2012^[Etienne, R.S., Haegeman, B., Stadler, T., Aze, T., Pearson, P.N., Purvis, A., Phillimore, A.B.: Diversity-dependence brings molecular phylogenies closer to agreement with the
fossil record. Proceedings of the Royal Society of London B: Biological Sciences
279(1732), 1300–1309 (2012) ].
The main purpose of the package is to provide a tool to estimate, given a phylogenetic tree, the likelihood for a set of three parameters: $\lambda$,$\mu$ and $q$.
Here:

- $\lambda$ is the sympatric speciation rate;

- $\mu$ is the extinction rate;

- $\nu$ is the multiple allopatric speciation trigger rate;

- $q$ is the single-lineage speciation probability.

Those parameters may be inferred through a maximum likelihood approach using the function mbd_ML.

## MBD usage

First clear memory and load the package:

```{r}
rm(list = ls()); options(warn = -1);
suppressWarnings( suppressPackageStartupMessages( library(MBD) ) )
```

Then we set some basic parameters to create a simulated tree:

```{r}
set.seed(14) #set.seed(42)
sim_pars <- c(0.2, 0.15, 2, 0.1); soc <- 2; cond <- 1; age <- 10;#sim_pars <- c(0.2, 0.1, 1.5, 0.1); soc = 2; cond = 1; age = 10;
sim <- MBD:::mbd_sim(pars = sim_pars, soc = soc, age = age, cond = cond, tips_interval = c(0, Inf))

```

Here "sim_pars" are $\lambda$, $\mu$, $\nu$ and $q$, i.e. the parameters we want to eventually be able to infer. There are also additional auxilary parameters such as "soc" (equal to 1 if we want a stem, or 2 if we want a crown), "cond" (let us condition the tree on the survival of crown species) and "age" (being the total age of the tree).

The "sim" list contain different objects.
You can visualize the full tree and keep track of species that die before present time (missing species):

```{r, fig.show='hold'}
plot(sim$tas)
```

If you want you can also show the reconstrucred tree, obtained cutting the kinks from the previous tree:

```{r, fig.show='hold'}
plot(sim$tes)
```

The dataframe containing all the information related to this tree is labeled as L (or L-table). You can access it by typing:

```{r}
sim$L
```

In this table:

- the first column are the times at which species are born;

- the second contains labels of the species' parents; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the third contains the labels of the daughter species themselves; positive and negative values only indicate whether the species belongs to the left or right crown lineage;

- the fourth contains the extinction times of the species. If this is equal to -1, then the species is still extant.

To recover the vector of branching times use:

```{r}
sim$brts
```

In this vector values repeated more than once indicate that a multiple birth process has taken place.
Since we are not interested in the topology of the tree, this vector alone can be used to calculate the likelihood of the tree with the function "mbd_loglik":

```{r}
test_pars <- c(0.3, 0.05, 1.0, 0.08)
MBD::mbd_loglik(pars = test_pars, brts = sim$brts, soc = soc, cond = cond, missnumspec = 0)
```

You can also ask for an optimization of the likelihood function using the function "mbd_ML" either considering a variation of all the parameters ($\lambda, \mu, \nu, q$) or any their subset of your choice. If you want for example to maximize the likelihood only for the parameter q:

```{r}
# # Uncomment if you are keen to wait.
# idparsopt <- 4
# ids <- 1:4; idparsfix <- ids[-idparsopt];
# parsfix <- sim_pars[idparsfix]; initparsopt <- 0.15;
# MBD:::mbd_ML(brts = sim$brts, 
#              initparsopt = initparsopt, 
#              idparsopt = idparsopt, 
#              parsfix = parsfix, 
#              idparsfix = idparsfix, 
#              soc = soc, 
#              cond = cond)

```
